### 2013-07-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
finished add step problem to nonlinear chapter

%%---------------------------------------------------------------------------%%
\section{Navier-Stokes Benchmark Problems\ }
\label{sec:ns_benchmarks}
To verify the FANM method for nonlinear problems, we choose benchmark
solutions for the 2-dimensional, steady, incompressible Navier-Stokes
equations on a rectilinear grid in much the same way as Shadid and
Pawlowski's work on Newton-Krylov methods for the solution of these
equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
define these equations as follows:
\begin{subequations}
  \begin{gather}
    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
    \ve{g} = \ve{0}
    \label{eq:ns_momentum}\\
    \nabla \cdot \ve{u} = 0
    \label{eq:ns_continuity}\\
    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
    \label{eq:ns_energy}
  \end{gather}
  \label{eq:navier_stokes}
\end{subequations}
where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
$C_p$ the specific heat capacity at constant pressure of the fluid,
$T$ the temperature of the fluid, and $\ve{g}$ the acceleration due to
gravity. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
dissipation effects neglected. In addition, we close the system with
the following equations:
\begin{subequations}
  \begin{gather}
    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
    \label{eq:ns_stress_tensor}\\
    \ve{q} = - k \nabla T\:,
    \label{eq:ns_heat_flux}
  \end{gather}
  \label{eq:ns_closure}
\end{subequations}
where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
in the fluid, and $k$ is the thermal conductivity of the fluid. This
set of strongly coupled equations possesses both the nonlinearities
and asymmetries that we are seeking for qualification of the FANM
method. Further, physical parameters within these equations can be
tuned to enhance the nonlinearities. We will then apply these
equations to the following three standard benchmark problems that
cover convection dominated flow and flow dominated by intertial
forces in two different geometries.

@@

\subsection{Lid Driven Cavity Problem}
\label{subsec:lid_driven_cavity}
To study performance for flow driven by intertial forces, the second
benchmark problem given by Ghia \citep{ghia_high-re_1982} incorporates
a driver for the flow to introduce higher Reynolds numbers into the
system, providing more inertial force to overcome the viscous forces
in the fluid. The setup for this problem is equally simple, containing
only the Dirichlet conditions as given in
Figure~\ref{fig:lid_driven_cavity} and is only applied to the mass and
momentum equations on the unit square.

@@

Isocurves of the velocity magnitudes are given in
Figure~\ref{fig:driven_velocity_isocurves} for solutions to the lid
driven cavity problem at Reynolds numbers of 100, 300, 500 and
700. Increasing the velocity on the top boundary of the problem drives
up the Reynolds number of the system making it more difficult to
solve. More rotation of the fluid is induced by inertial forces and we
begin to see some additional vortices form besides the primary vortex
near the center of the box.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_velocity_isocurves.png}
  \end{center}
  \caption{\textbf{Velocity magnitude isocurves from solution to the
      backward facing step problem.} \textit{From top to bottom: Re =
      200, Re = 300, Re = 400, Re = 500. Isocurves on unit velocity
      scale from 0 to 1 with 20 uniformly spaced curves for each
      calculation.}}
  \label{fig:step_velocity_isocurves}
\end{figure}

@@

For every benchmark calculation, both FANM and Newton-Krylov were
preconditioned with the same multigrid parameters such that the
conditioning of the system would not be a factor in the verification
of the method or the performance comparison in the following
section. With FANM, this preconditioning was required to reduce the
spectral radius of the Jacobian operators generated at each Newton
step. This preconditioning was required to obtain good convergence
properties for the Newton-Krylov solutions although it was not
necessary for their convergence. In addition, the same nonlinear
solver parameters including globalization and forcing term strategies
were used with both methods. As with the radiation transport
calculations in the previous chapter, algrebraic preconditioning with
the multigrid method results in dense linear systems. To mitigate
this, the reduced domain approximation was again applied to the Monte
Carlo solution within the MCSA solve occuring at each nonlinear
iteration. The values used for the approximation will be given for
each benchmark problem.

Results in this and the following section were produced by the Drekar
multiphysics code base being developed at Sandia National Laboratories
for fluid flow and other multiple physics simulations
\citep{pawlowski_drekar_2012}. Drekar is a massively parallel physics
framework that utilizes Newton-Krylov methods to generate a fully
consistent solution scheme for the Navier-Stokes equations among other
systems with Jacobian generation through automatic differentiation. To
achieve these results, a Monte Carlo solver generated for this work
was injected into the nonlinear toolchain within Drekar to form an
implementation of the FANM method. For the Newton-Krylov solutions,
the same production implementation of GMRES from the Trilinos
scientific computing library Aztec \citep{heroux_overview_2005} used
for the verification of MCSA for the $SP_N$ solutions in the previous
chapter was also used here as a means of comparison and FANM
benchmarking.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lcc}\hline\hline
      \multicolumn{1}{l}{Parameter}& 
      \multicolumn{1}{c}{Value}&
      \multicolumn{1}{c}{Units}\\\hline
      $x_{min}$ & 0.0 & $m$ \\
      $x_{max}$ & 1.0 & $m$ \\
      $y_{min}$ & 0.0 & $m$ \\
      $y_{max}$ & 1.0 & $m$ \\
      $N_x$ & 50 & - \\
      $N_y$ & 50 & - \\
      $DOFs$ & 10404 & - \\
      Initial Temperature & 0.0 & $K$ \\
      Density & 1.0 & $kg / m^3$ \\
      Specific Heat & 1.0 & J / $(kg \times K)$ \\
      Dynamic Viscosity & 0.71 & $(N \times s) / m^2$ \\
      Thermal Conductivity & 1.0 & $W / (m \times K)$ \\
      Thermal Expansion Coefficient & 1000 & $1 / K$ \\
      Thermal Diffusivity & 1.0 & $m^2 / s$ \\
      Kinematic Viscosity & 0.71 & $m^2 / s$ \\
      Gravitational Acceleration & 10 & $m / s^2$ \\
      Cavity Characteristic Length & 1.0 & $m$ \\
      Prandtl Number & 0.71 & - \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Thermal convection cavity FANM verification model
      problem parameters.}  \textit{The Navier-Stokes equations in 2
      dimensions are used for the model problem. The Rayleigh number
      is varied by modifying the hot boundary temperature to induce
      larger temperature gradients and bouyancy-driven flow.}}
  \label{tab:thermal_convection_parameters}
\end{table}

@@

In the literature on this benchmark, solutions at Reynolds numbers of
up to \sn{1}{4} were achieved using a variety of nonlinear solution
techniques. With the FANM method, at Reynolds numbers above 700
convergence of the linear iterations could not be achieved regardless
of how agressive the multigrid preconditioner was set. In these cases,
the Jacobian operators are ill-conditioned enough that even GMRES
converges slowly relative to the thermal convection cavity
benchmark. Therefore, the spectral radius limitation on MCSA is
extremely prohibitive in this case and others where transport driven
by inertial forces is dominant. Because of this, only solutions to
the lid driven cavity benchmark at Reynolds numbers of up to 700 are
presented.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lc}\hline\hline
      \multicolumn{1}{l}{Parameter}& 
      \multicolumn{1}{c}{Value}\\\hline
      Weight Cutoff & \sn{1}{-2} \\
      Fixed Point Iteration & Richardson \\
      Estimator & Adjoint Collision \\
      Histories, Re=200 & 500,000 per iteration \\
      Histories, Re=300 & 500,000 per iteration \\
      Histories, Re=400 & 500,000 per iteration \\
      Histories, Re=500 & 1,000,000 per iteration \\
      Reduced Domain Fill Level, All Re & 200 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step FANM verification MCSA solver
      parameters.} \textit{No relaxation parameters or variance
      reduction techniques were used with MCSA other than the reduced
      domain approximation. The same fill level for the reduced domain
      approximation was used at every Reynold numbers and different
      histories per iteration were used at different Reynolds numbers
      due to convergence requirements.}}
  \label{tab:step_mcsa_parameters}
\end{table}

In the literature on this benchmark, solutions at Reynolds numbers of
up to 800 were achieved using a variety of nonlinear solution
techniques. In identical fashion to the lid driven cavity problem, at
Reynolds numbers above 500 convergence of the linear iterations could
not be achieved with FANM regardless of how agressive the multigrid
preconditioner was set. In these cases, the Jacobian operators are
again ill-conditioned enough that a spectral radius of less than one
could not be achieved with multigrid preconditioning and several cycle
applications. Therefore, the spectral radius limitation on MCSA is
extremely prohibitive in this case as well and even more so than the
lid driven cavity problem where solutions at higher Reynolds numbers
were achieved. Because of this, only solutions to the backward facing
step benchmark at Reynolds numbers of up to 500 are presented.

@@

At each Rayleigh number used for the thermal convection cavity
problem, Table~\ref{tab:convection_nonlinear_iter_comparison} gives
the number of Newton iterations required for convergence for each
method.  For all Rayleigh numbers except \sn{1}{5} the number of
Newton iterations required to converge is the same while the \sn{1}{5}
result required an extra FANM iteration. This is due to the fact that
the convergence of the linear residual has a strong effect on the
convergence of the nonlinear residual. For nonlinear calculations,
this is much stronger than the effect on the convergence of the
eigenvalue in the $SP_N$ computations where we always observed that
identical numbers of k-eigenvalue iterations were required to converge
each variation of the probem.

@@

For the \sn{1}{5} Rayleigh number case, this means that the GMRES
solver at each Newton step is converging the linear residual to a
smaller tolerance than the MCSA solver as the forcing terms adapt to
the convergence of the method over the course of nonlinear
iterations. As the Rayleigh number increases, the problem becomes more
difficult to solve and therefore a more agressive preconditioner
should maintain iterative performance of the nonlinear method. This
was observed to be true for the \sn{1}{6} Rayleigh number case. For
this calculation, the number of multigrid cycle applications was
bumped up to 3 from 1 and FANM converged in the same number of
nonlinear iterations as the Newton-Krylov method.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/ra_1e5_convergence.pdf}
  \end{center}
  \caption{\textbf{Nonlinear residual norm as a function of Newton
      iteration for the \sn{1}{5} Rayleigh number case.} \textit{The
      FANM calculation required an additional nonlinear iteration to
      converge.}}
  \label{fig:ra1e5_convergence}
\end{figure}

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/convection_ra1e6_iters.pdf}
  \end{center}
  \caption{\textbf{Linear solver iterations as a function of Newton
      iteration for the \sn{1}{6} Rayleigh number case.} \textit{MCSA
      shows better iterative performance than GMRES at all but the
      second nonlinear iteration. The forcing term was fixed at
      \sn{1}{-4} at all nonlinear iterations.}}
  \label{fig:ra1e6_linear_iters}
\end{figure}

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/re700_convergence.pdf}
  \end{center}
  \caption{\textbf{Nonlinear residual norm as a function of Newton
      iteration for the 700 Reynolds number case.} \textit{The FANM
      calculation converged in 4 fewer iterations than the
      Newton-Krylov calculation.}}
  \label{fig:re700_convergence}
\end{figure}

@@

As an additional means of analysis for the linear solver iteration
results, we will look at a case where MCSA performed poorly compared
to GMRES. Consider the linear iterations as a function of nonlinear
iteration given in Figure~\ref{fig:driven_re500_iters} and the forcing
term as a function of nonlinear iteration given in
Figure~\ref{fig:driven_re500_forcing} for the 500 Reynolds number
case. At the later Newton iterations where the region of fast
quadratic convergence is reached, we see that MCSA required many more
iterations to converge than GMRES, accounting for the majority of the
extra iterations. However, we see in the forcing term data that
forcing terms at later Newton iterations for MCSA were much smaller
than those for GMRES. This means that MCSA was converged to a tighter
tolerance at these iterations and therfore required the extra
iterations.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_iters.pdf}
  \end{center}
  \caption{\textbf{Linear solver iterations as a function of Newton
      iteration for the 500 Reynolds number case.} \textit{More MCSA
      iterations were required to converge the linear models at later
      Newton iterations due to smaller forcing terms.}}
  \label{fig:driven_re500_iters}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_forcing.pdf}
  \end{center}
  \caption{\textbf{Forcing terms as a function of Newton
      iteration for the 500 Reynolds number case.} \textit{Once the
      region of fast quadratic convergence is reached, MCSA forcing
      terms were much larger.}}
  \label{fig:driven_re500_forcing}
\end{figure}

@@

As a final comparison, the timing of both methods was again compared
with the speedup of using the Newton-Krylov solver over FANM reported
in Table~\ref{tab:driven_speedup_comparison}. The speedup values
reported here are of the same magnitude as those reported for the
thermal convection cavity problem and again in agreement with those
observed for the $SP_N$ calculations. Lack of signficant variation in
these timing results also point to a qualititively similar time
complexity for the FANM method as a function of primary problem
parameter relative to the Newton-Krylov method. In all cases, the
preconditioning strategy must be signficantly improved along with
implementation optimization to avoid these large overheads.

@@

Unlike the lid driven cavity benchmark, the extra linear solver
iterations are not due to the smaller forcing terms but are rather
truly a function of the ill-conditioning of the Jacobian matrices
produced at each Newton step.

@@

As a final overall comparison, we tabulate the results from the
benchmark calculations over the three performance metrics: nonlinear
iterations, linear iterations, and CPU
time. Tables~\ref{tab:benchmark_nonlinear_comparison},
\ref{tab:benchmark_linear_comparison}, and
\ref{tab:benchmark_time_comparison} give the results. In these tables,
the calculations that procduced the best performing results are
tallied. In the case of a tie, both are tallied. Looking at the
tabulated results, we see marginally better performance for FANM in
terms of nonlinear iterations with 1 extra case outperforming the
Newton-Krylov method. As observed for each benchmark, the timing
results were always in favor of the Newton-Krylov method due to the
effects of the explict preconditioning strategy.

For the total linear solver iterations given in
Table~\ref{tab:benchmark_linear_comparison}, the difference is more
dramatic with FANM converging in fewer the nonlinear residual in fewer
linear solver iterations for twice as many cases as the Newton-Krylov
method keeping in mind that for all cases both GMRES and MCSA were
preconditioned with identical multigrid parameters using the same
multigrid preconditioning library. This shows that when properly
preconditioned, we can expect performance improvements in terms of
total linear solver iterations required to converge the nonliner
residual. With future optimization of MCSA implementations, this may
also mean improved time to solution for FANM when compared to
Newton-Krylov methods.

@@

In this chapter we have developed and explored Monte Carlo synthetic
acceleration methods in the context of solutions to the nonlinear
Navier-Stokes equations. The following are the signficant observations
and findings.

>>>>

%%---------------------------------------------------------------------------%%
\section{Navier-Stokes Benchmark Problems\ }
\label{sec:ns_benchmarks}
To verify the FANM method for nonlinear problems, we choose benchmark
solutions for the 2-dimensional, steady, incompressible Navier-Stokes
equations on a rectilinear grid in much the same way as Shadid and
Pawlowski's work on Newton-Krylov methods for the solution of these
equations \citep{shadid_inexact_1997,pawlowski_globalization_2006}. We
define these equations as follows:
\begin{subequations}
  \begin{gather}
    \rho \ve{u} \cdot \nabla \ve{u} - \nabla \cdot \ve{T} - \rho
    \ve{g} = \ve{0}
    \label{eq:ns_momentum}\\
    \nabla \cdot \ve{u} = 0
    \label{eq:ns_continuity}\\
    \rho C_p \ve{u} \cdot \nabla T + \nabla \cdot \ve{q} = 0\:,
    \label{eq:ns_energy}
  \end{gather}
  \label{eq:navier_stokes}
\end{subequations}
where $\rho$ is the fluid density, $\ve{u}$ is the fluid velocity,
$C_p$ the specific heat capacity at constant pressure of the fluid,
$T$ the temperature of the fluid, and $\ve{g}$ the acceleration due to
gravity. Eq~(\ref{eq:ns_momentum}) provides momentum transport,
Eq~(\ref{eq:ns_continuity}) provides the mass balance, and
Eq~(\ref{eq:ns_energy}) provides energy transport with viscous
dissipation effects neglected. In addition, we close the system with
the following equations:
\begin{subequations}
  \begin{gather}
    \ve{T} = -P \ve{I} + \mu[\nabla \ve{u} + \nabla \ve{u}^T]
    \label{eq:ns_stress_tensor}\\
    \ve{q} = - k \nabla T\:,
    \label{eq:ns_heat_flux}
  \end{gather}
  \label{eq:ns_closure}
\end{subequations}
where $\ve{T}$ is the stress tensor, $P$ is the hydrodynamic pressure,
$\mu$ is the dynamic viscosity of the fluid, $\ve{q}$ is the heat flux
in the fluid, and $k$ is the thermal conductivity of the fluid. This
set of strongly coupled equations possesses both the nonlinearities
and asymmetries that we are seeking for qualification of the FANM
method. Further, physical parameters within these equations can be
tuned to enhance the nonlinearities. We will then apply these
equations to the following three standard benchmark problems that
cover convection dominated flow and flow dominated by inertial
forces in two different geometries.

@@

\subsection{Lid Driven Cavity Problem}
\label{subsec:lid_driven_cavity}
To study performance for flow driven by inertial forces, the second
benchmark problem given by Ghia \citep{ghia_high-re_1982} incorporates
a driver for the flow to introduce higher Reynolds numbers into the
system, providing more inertial force to overcome the viscous forces
in the fluid. The setup for this problem is equally simple, containing
only the Dirichlet conditions as given in
Figure~\ref{fig:lid_driven_cavity} and is only applied to the mass and
momentum equations on the unit square.

@@

Isocurves of the velocity magnitudes are given in
Figure~\ref{fig:driven_velocity_isocurves} for solutions to the lid
driven cavity problem at Reynolds numbers of 100, 300, 500 and
700. Increasing the velocity on the top boundary of the problem drives
up the Reynolds number of the system making it more difficult to
solve. More rotation of the fluid is induced by inertial forces and we
begin to see some additional vortexes form besides the primary vortex
near the center of the box.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/step_velocity_isocurves.png}
  \end{center}
  \caption{\textbf{Velocity magnitude isocurves from solution to the
      backward facing step problem.} \textit{From top to bottom: Re =
      200, Re = 300, Re = 400, Re = 500. Isocurves on unit velocity
      scale from 0 to 1 with 20 uniformly spaced curves for each
      calculation. The depression forming in the upper right portion
      of visible flow is the second vortex.}}
  \label{fig:step_velocity_isocurves}
\end{figure}

@@

For every benchmark calculation, both FANM and Newton-Krylov were
preconditioned with the same multigrid parameters such that the
conditioning of the system would not be a factor in the verification
of the method or the performance comparison in the following
section. With FANM, this preconditioning was required to reduce the
spectral radius of the Jacobian operators generated at each Newton
step. This preconditioning was required to obtain good convergence
properties for the Newton-Krylov solutions although it was not
necessary for their convergence. In addition, the same nonlinear
solver parameters including globalization and forcing term strategies
were used with both methods. As with the radiation transport
calculations in the previous chapter, algebraic preconditioning with
the multigrid method results in dense linear systems. To mitigate
this, the reduced domain approximation was again applied to the Monte
Carlo solution within the MCSA solve occurring at each nonlinear
iteration. The values used for the approximation will be given for
each benchmark problem.

Results in this and the following section were produced by the Drekar
multiphysics code base being developed at Sandia National Laboratories
for fluid flow and other multiple physics simulations
\citep{pawlowski_drekar_2012}. Drekar is a massively parallel physics
framework that utilizes Newton-Krylov methods to generate a fully
consistent solution scheme for the Navier-Stokes equations among other
systems with Jacobian generation through automatic differentiation. To
achieve these results, a Monte Carlo solver generated for this work
was injected into the nonlinear tool chain within Drekar to form an
implementation of the FANM method. For the Newton-Krylov solutions,
the same production implementation of GMRES from the Trilinos
scientific computing library Aztec \citep{heroux_overview_2005} used
for the verification of MCSA for the $SP_N$ solutions in the previous
chapter was also used here as a means of comparison and FANM
benchmarking.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lcc}\hline\hline
      \multicolumn{1}{l}{Parameter}& 
      \multicolumn{1}{c}{Value}&
      \multicolumn{1}{c}{Units}\\\hline
      $x_{min}$ & 0.0 & $m$ \\
      $x_{max}$ & 1.0 & $m$ \\
      $y_{min}$ & 0.0 & $m$ \\
      $y_{max}$ & 1.0 & $m$ \\
      $N_x$ & 50 & - \\
      $N_y$ & 50 & - \\
      $DOFs$ & 10404 & - \\
      Initial Temperature & 0.0 & $K$ \\
      Density & 1.0 & $kg / m^3$ \\
      Specific Heat & 1.0 & J / $(kg \times K)$ \\
      Dynamic Viscosity & 0.71 & $(N \times s) / m^2$ \\
      Thermal Conductivity & 1.0 & $W / (m \times K)$ \\
      Thermal Expansion Coefficient & 1000 & $1 / K$ \\
      Thermal Diffusivity & 1.0 & $m^2 / s$ \\
      Kinematic Viscosity & 0.71 & $m^2 / s$ \\
      Gravitational Acceleration & 10 & $m / s^2$ \\
      Cavity Characteristic Length & 1.0 & $m$ \\
      Prandtl Number & 0.71 & - \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Thermal convection cavity FANM verification model
      problem parameters.}  \textit{The Navier-Stokes equations in 2
      dimensions are used for the model problem. The Rayleigh number
      is varied by modifying the hot boundary temperature to induce
      larger temperature gradients and buoyancy-driven flow.}}
  \label{tab:thermal_convection_parameters}
\end{table}

@@

In the literature on this benchmark, solutions at Reynolds numbers of
up to \sn{1}{4} were achieved using a variety of nonlinear solution
techniques. With the FANM method, at Reynolds numbers above 700
convergence of the linear iterations could not be achieved regardless
of how aggressive the multigrid preconditioner was set. In these cases,
the Jacobian operators are ill-conditioned enough that even GMRES
converges slowly relative to the thermal convection cavity
benchmark. Therefore, the spectral radius limitation on MCSA is
extremely prohibitive in this case and others where transport driven
by inertial forces is dominant. Because of this, only solutions to
the lid driven cavity benchmark at Reynolds numbers of up to 700 are
presented.

@@

\begin{table}[h!]
  \begin{center}
    \begin{tabular}{lc}\hline\hline
      \multicolumn{1}{l}{Parameter}& 
      \multicolumn{1}{c}{Value}\\\hline
      Weight Cutoff & \sn{1}{-2} \\
      Fixed Point Iteration & Richardson \\
      Estimator & Adjoint Collision \\
      Histories, Re=200 & 500,000 per iteration \\
      Histories, Re=300 & 500,000 per iteration \\
      Histories, Re=400 & 500,000 per iteration \\
      Histories, Re=500 & 1,000,000 per iteration \\
      Reduced Domain Fill Level, All Re & 200 \\
      %%
      \hline\hline
    \end{tabular}
  \end{center}
  \caption{\textbf{Backward facing step FANM verification MCSA solver
      parameters.} \textit{No relaxation parameters or variance
      reduction techniques were used with MCSA other than the reduced
      domain approximation. The same fill level for the reduced domain
      approximation was used at every Reynolds numbers and different
      histories per iteration were used at different Reynolds numbers
      due to convergence requirements.}}
  \label{tab:step_mcsa_parameters}
\end{table}

In the literature on this benchmark, solutions at Reynolds numbers of
up to 800 were achieved using a variety of nonlinear solution
techniques. In identical fashion to the lid driven cavity problem, at
Reynolds numbers above 500 convergence of the linear iterations could
not be achieved with FANM regardless of how aggressive the multigrid
preconditioner was set. In these cases, the Jacobian operators are
again ill-conditioned enough that a spectral radius of less than one
could not be achieved with multigrid preconditioning and several cycle
applications. Therefore, the spectral radius limitation on MCSA is
extremely prohibitive in this case as well and even more so than the
lid driven cavity problem where solutions at higher Reynolds numbers
were achieved. Because of this, only solutions to the backward facing
step benchmark at Reynolds numbers of up to 500 are presented.

@@

At each Rayleigh number used for the thermal convection cavity
problem, Table~\ref{tab:convection_nonlinear_iter_comparison} gives
the number of Newton iterations required for convergence for each
method.  For all Rayleigh numbers except \sn{1}{5} the number of
Newton iterations required to converge is the same while the \sn{1}{5}
result required an extra FANM iteration. This is due to the fact that
the convergence of the linear residual has a strong effect on the
convergence of the nonlinear residual. For nonlinear calculations,
this is much stronger than the effect on the convergence of the
eigenvalue in the $SP_N$ computations where we always observed that
identical numbers of k-eigenvalue iterations were required to converge
each variation of the problem.

@@

For the \sn{1}{5} Rayleigh number case, this means that the GMRES
solver at each Newton step is converging the linear residual to a
smaller tolerance than the MCSA solver as the forcing terms adapt to
the convergence of the method over the course of nonlinear
iterations. As the Rayleigh number increases, the problem becomes more
difficult to solve and therefore a more aggressive preconditioner
should maintain iterative performance of the nonlinear method. This
was observed to be true for the \sn{1}{6} Rayleigh number case. For
this calculation, the number of multigrid cycle applications was
bumped up to 3 from 1 and FANM converged in the same number of
nonlinear iterations as the Newton-Krylov method.

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/ra_1e5_convergence.pdf}
  \end{center}
  \caption{\textbf{Thermal convection cavity nonlinear residual norm
      as a function of Newton iteration for the \sn{1}{5} Rayleigh
      number case.} \textit{The FANM calculation required an
      additional nonlinear iteration to converge.}}
  \label{fig:ra1e5_convergence}
\end{figure}

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/convection_ra1e6_iters.pdf}
  \end{center}
  \caption{\textbf{Thermal convection cavity linear solver iterations
      as a function of Newton iteration for the \sn{1}{6} Rayleigh
      number case.} \textit{MCSA shows better iterative performance
      than GMRES at all but the second nonlinear iteration. The
      forcing term was fixed at \sn{1}{-4} at all nonlinear
      iterations.}}
  \label{fig:ra1e6_linear_iters}
\end{figure}

@@

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/re700_convergence.pdf}
  \end{center}
  \caption{\textbf{Lid driven cavity nonlinear residual norm as a
      function of Newton iteration for the 700 Reynolds number case.}
    \textit{The FANM calculation converged in 4 fewer iterations than
      the Newton-Krylov calculation.}}
  \label{fig:re700_convergence}
\end{figure}

@@

As an additional means of analysis for the linear solver iteration
results, we will look at a case where MCSA performed poorly compared
to GMRES. Consider the linear iterations as a function of nonlinear
iteration given in Figure~\ref{fig:driven_re500_iters} and the forcing
term as a function of nonlinear iteration given in
Figure~\ref{fig:driven_re500_forcing} for the 500 Reynolds number
case. At the later Newton iterations where the region of fast
quadratic convergence is reached, we see that MCSA required many more
iterations to converge than GMRES, accounting for the majority of the
extra iterations. However, we see in the forcing term data that
forcing terms at later Newton iterations for MCSA were much smaller
than those for GMRES. This means that MCSA was converged to a tighter
tolerance at these iterations and therefore required the extra
iterations.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_iters.pdf}
  \end{center}
  \caption{\textbf{Lid driven cavity linear solver iterations as a
      function of Newton iteration for the 500 Reynolds number case.}
    \textit{More MCSA iterations were required to converge the linear
      models at later Newton iterations due to smaller forcing
      terms.}}
  \label{fig:driven_re500_iters}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/driven_re500_forcing.pdf}
  \end{center}
  \caption{\textbf{Lid driven cavity forcing terms as a function of
      Newton iteration for the 500 Reynolds number case.} \textit{Once
      the region of fast quadratic convergence is reached, MCSA
      forcing terms were much larger.}}
  \label{fig:driven_re500_forcing}
\end{figure}

@@

As a final comparison, the timing of both methods was again compared
with the speedup of using the Newton-Krylov solver over FANM reported
in Table~\ref{tab:driven_speedup_comparison}. The speedup values
reported here are of the same magnitude as those reported for the
thermal convection cavity problem and again in agreement with those
observed for the $SP_N$ calculations. Lack of significant variation in
these timing results also point to a qualitatively similar time
complexity for the FANM method as a function of primary problem
parameter relative to the Newton-Krylov method. In all cases, the
preconditioning strategy must be significantly improved along with
implementation optimization to avoid these large overheads.

@@

Unlike the lid driven cavity benchmark, the extra linear solver
iterations are not due to the smaller forcing terms but are rather
truly a function of the ill-conditioning of the Jacobian matrices
produced at each Newton step. We see this by comparing the results at
Reynolds numbers of 400 where FANM outperformed Newton-Krylov and 500
where Newton-Krylov outperformed FANM. Linear iterations as a function
of nonlinear iteration given in Figure~\ref{fig:step_re400_iters} and
the forcing term as a function of nonlinear iteration given in
Figure~\ref{fig:step_re400_forcing} for the 400 Reynolds number case
and the same data is given in Figure~\ref{fig:step_re500_iters} and
Figure~\ref{fig:step_re500_forcing} for the 500 Reynolds number
case. For both cases, FANM always has larger forcing terms and should
therefore use fewer MCSA iterations at each Newton step to converge
the linear model. This is indeed the case for the 400 Reynolds number
case where the linear model is better conditioned and FANM used fewer
total linear solver iterations to converge the nonlinear
residual. However, at a Reynolds number of 500 FANM has larger forcing
terms but requires many more MCSA iterations to converge the last 2
Newton steps. Therefore, these extra iterations in this case must come
from ill-conditioning.

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/step_re400_iters.pdf}
  \end{center}
  \caption{\textbf{Backward facing step linear solver iterations as a
      function of Newton iteration for the 400 Reynolds number case.}
    \textit{More MCSA iterations were required to converge the linear
      models at later Newton iterations due to ill-conditioning of the
      Jacobian matrices.}}
  \label{fig:step_re400_iters}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/step_re400_forcing.pdf}
  \end{center}
  \caption{\textbf{Backward facing step forcing terms as a function of
      Newton iteration for the 400 Reynolds number case.} \textit{FANM
      forcing terms were always larger.}}
  \label{fig:step_re400_forcing}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/step_re500_iters.pdf}
  \end{center}
  \caption{\textbf{Backward facing step linear solver iterations as a
      function of Newton iteration for the 500 Reynolds number case.}
    \textit{More MCSA iterations were required to converge the linear
      models at later Newton iterations due to ill-conditioning of the
      Jacobian matrices.}}
  \label{fig:step_re500_iters}
\end{figure}

\begin{figure}[t!]
  \begin{center}
    \includegraphics[width=6in]{chapters/nonlinear_problem/step_re500_forcing.pdf}
  \end{center}
  \caption{\textbf{Backward facing step forcing terms as a function of
      Newton iteration for the 500 Reynolds number case.} \textit{More
      MCSA iterations were required to converge the linear models at
      later Newton iterations due to ill-conditioning of the Jacobian
      matrices.}}
  \label{fig:step_re500_forcing}
\end{figure}

The ill-conditioning of the 500 Reynolds number case is also apparent
in the timing data given in
Table~\ref{tab:step_speedup_comparison}. For the first two cases, we
see qualitatively similar speedup values for the Newton-Krylov solver
relative to the observations for the first two benchmark problems of
O(100). However, the we expect the ill-conditioning to cause the Monte
Carlo histories in the MCSA solver to have longer random walk lengths
based on our MCSA breakdown analysis in the previous chapter. In
addition, we observed that the linear model required many more MCSA
iterations to converge. These longer random walks and more iterations
manifest themselves in the longer CPU time values and therefore the
larger Newton Krylov speedup value of O(1,000) for the 500 Reynolds
number case.

@@

As a final overall comparison, we tabulate the results from the
benchmark calculations over the three performance metrics: nonlinear
iterations, linear iterations, and CPU
time. Tables~\ref{tab:benchmark_nonlinear_comparison},
\ref{tab:benchmark_linear_comparison}, and
\ref{tab:benchmark_time_comparison} give the results. In these tables,
the calculations that produced the best performing results are
tallied. In the case of a tie, both are tallied. Looking at the
tabulated results, we see marginally better performance for FANM in
terms of nonlinear iterations with 1 extra case outperforming the
Newton-Krylov method. As observed for each benchmark, the timing
results were always in favor of the Newton-Krylov method due to the
effects of the explicit preconditioning strategy.

For the total linear solver iterations given in
Table~\ref{tab:benchmark_linear_comparison}, the difference is more
dramatic with FANM converging in fewer the nonlinear residual in fewer
linear solver iterations for twice as many cases as the Newton-Krylov
method keeping in mind that for all cases both GMRES and MCSA were
preconditioned with identical multigrid parameters using the same
multigrid preconditioning library. This shows that when properly
preconditioned, we can expect performance improvements in terms of
total linear solver iterations required to converge the nonlinear
residual. With future optimization of MCSA implementations, this may
also mean improved time to solution for FANM when compared to
Newton-Krylov methods.

@@

In this chapter we have developed and explored Monte Carlo synthetic
acceleration methods in the context of solutions to the Navier-Stokes
equations. The following are the significant observations and findings.
